Parallel air palletization and tour planning for simultaneous pickup and delivery

### 2023-02-09 latex/comp_or.tex
Running data50 solutions

\cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.

@@

An important issue on meta-heuristics is that they are known for the hard work in tuning parameters. For this, when we use the word {\it empirically}\/ throughout the text, we want to say that we accomplished many tests with the hardest instances to fit these values. This does not guarantee best results for all instances, but, on average, overall results are acceptable. Another important observation is that our {\it ACO} implementations were designed based on traditional ideas from the literature \cite{Fidanova2006} and on the experiences learned during tests, plus the special features to process-based parallelism.

@@

The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$. It is directly proportional to the score, and inversely proportional to the volume and the torque of each item. The 3000 value was empirically set to keep the attractiveness in the range [0, 1], to be in the same order of magnitude of the Ant Colony Optimization pheromone range.

@@

Below, we present 2 implementations for $SolveNode$: {\it Multi-process Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.

@@

It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel (for a detailed information on parallel ACO see \cite{Pedemonte2011}). We chose {\it mpACO} for our approach because we employed the {\it Fully-Connected} topology which requires to share the pheromone array between the multi-process ants and the facility to handle the possibility of race condition.

\subsubsection{Multi-process Ant Colony Optimization (mpACO)}

In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search (sequential or in parallel), where each ant $a$\/ finds its own allocation graph $G^a$. Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants. If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$.

Algorithm \ref{alg:aco} presents this heuristic in its parallel multi-process version as we have implemented.

$SolveNode(mpACO, ...)$ is the main process, while $AntSolve(...)$ is executed the child concurrent processes.

\begin{algorithm}[H]

	\caption{ Solves a node with the {\it mpACO} method } \label{alg:aco}
	\begin{algorithmic}[1]

		\Procedure{$SolveNode$}{$mpACO, k, G, numProcs,\ timeLim$}

		\State Let $G(V_k, E^Q_k)$ \Comment{This is a partial solution with only consolidated on board}
		\State $G^*(V_k, E^{N*}_k \cup E^Q_k), \_ \gets Greedy(k, G, 0.9)$  \label{aco:greedy} \Comment{This partial solution is completed by a greedy algorithm}
		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
		\State $A \gets numProcs$ \label{aco:numProcs}
		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
		\State $\tau^k_{ij} \gets 1$, $1 \leq i \leq m$, $1 \leq j \leq n_k$   \label{aco:init2}
		\State $\alpha    \gets 1.0$ \label{aco:alpha}
		\State $\beta     \gets 4.0$ \label{aco:beta}
		\State $\rho      \gets 0.2$ \label{aco:rho}
		\State $stagnant \gets 0$
		\While{ $stagnant < 3$ } \label{aco:while}

			\State $AntsQueue \gets \{\ \}$ \label{aco:AntsQueue}

			\For{$1$ {\bf to} $A$} \label{aco:AntsSolve}	
				\State $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$
			\EndFor

			\State $counter \gets A$

			\While{$counter > 0$} \label{aco:while2}

				\State $G^{a} \gets \varnothing$

					\State $G^{a} \gets AntsQueue.Pop()$

					\If{$G^{a} \ne \varnothing$}				
						\Break
					\EndIf

				\If{$f_s(G^{a}) > f_s(G^{+})$}
					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$ 
				\EndIf

				\State $counter \gets counter - 1$

			\If{$f_s(G^+) > f_s(G^*)$} \label{aco:best1}
				\State $stagnant \gets 0$
				\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$		
			\Else
				\State $stagnant \gets stagnant + 1$ 
			\EndIf
			\For{all $\tau^k_{ij}$} 
				\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho2}
				\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{aco:phero2}
				\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}				
			\EndFor	
		\EndWhile	
		\Return $G^*(V_k,\ E^{N*}_k \cup E^Q_k)$

A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$. The argument 0.9 generates an initial and poor solution for ACO to improve. The second return value denoted by underscore is not used by ACO.

The number $A$\/ of ants was set to the number of parallel processes (line \ref{aco:numProcs}).

The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants. For further explanation, see \cite{DorigoManiezzoColorni1996}.

The number of iterations depends on the time limit or stagnation in improving results (line \ref{aco:while}).
At the beginning os each iteration (\ref{aco:AntsQueue}) a queue of ants ($AntsQueue$) is created to hold ants parallel solutions generated in the child processes.

The local solution $G^+$ is sent to A child processes (\ref{aco:AntsSolve}) almost simultaneously.
While there are any ant working on an solution or the run-time is still less then maximum (\ref{aco:while2}), it hangs in the $AntsQueue$ until there is a solution available.

If there is a global improvement (\ref{aco:best1}) the best so far solution ($G^*(V_k, E^{N*}_k \cup E^Q_k)$) is updated.
At the end of each iteration pheromone levels (\ref{aco:rho2}) are evaporated, than updated with the new calculated $\Delta\tau$ (line \ref{aco:phero2}), and according to Equation \ref{eq:delta_tau}, and the general attractiveness ( \ref{aco:ga2}) is also updated.

\begin{equation} \label{eq:delta_tau}
	\Delta\tau = \frac{f_s(G^{+}) - f_s(G^{*})}{A \times f_s(G^{*})}
\end{equation}

Many ACO applications update pheromone levels and the general attractiveness only when there is a solution improvement. Differently, this work updates in both cases, exploring also the negative learning approach proposed by \cite{math9040361}. We did not compare these performances because it is out of the scope of this work.

@@

\begin{algorithm}[H] 
	\caption{Each ant solves a node in parallel} \label{alg:AntSolve}
	\begin{algorithmic}[1]

		\Procedure{$AntSolve$}{$G^+,\ AntsQueue,\ ga^k_{ij})$}

		\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
		\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$ \label{AntSolve:neighborhood}
		\While{$|E| > 0$} \label{AntSolve:not_empty}

			\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{AntSolve:roulette}
			\State $E \gets E - \{e\}$		
			\If{$E^{Na}_k \cup \{e\}$ is feasible}
				\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{AntSolve:ant}
			\EndIf

%\For{$\tau^k_{ij}, \theta^k_{ij} \in E^{Na}_k$}
%\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{AntSolve:rho}
%\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{AntSolve:phero}  \Comment{$\Delta\tau \gets$ Equation \ref{eq:delta_tau}}
%\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{AntSolve:ga}		
%\EndFor

		\State $AntsQueue \gets AntsQueue \cup G^a(V_k,\ E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}

Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution $G^+(V_k, E^{N+}_k \cup E^Q_k)$ generated greedily to be completed and improved by {\it mpACO}.

Line \ref{AntSolve:neighborhood} expresses the creation of the neighborhood, the set of candidate edges.

While the neighborhood is not empty (line \ref{AntSolve:not_empty}), choose an edge (line \ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (line \ref{AntSolve:ant}).

Finally, put the solution in the queue (line \ref{AntSolve:put_in_queue}) to be retrieved by the main process.

%Note that $ga^k_{ij}$ is updated by each parallel ant, which may influence the solution of the others during run-time. This fully connected topology may increase run-time due to the overhead in communications handled by {\it Python API Multiprocessing} but it is expected that this will yield better solutions.

@@

%
%\begin{figure}[H]
%	\begin{center}
%		\begin{tikzpicture}[scale=0.8, samples=100]
%			\begin{axis}[
%				myplotstyle,
%				xlabel={\bf mpShims computation times (s)},
%				ylabel={\bf Score},
%				]
%				\addplot+[
%				only marks,	
%				nodes near coords = {\numProcs{}},
%				visualization depends on = {value \thisrow{np} \as \numProcs},						
%				]
%				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
%			\end{axis}
%			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
%		\end{tikzpicture}%
%		~%
%		%
%		\begin{tikzpicture}[scale=0.8, samples=100]
%		\begin{axis}[
%			myplotstyle,
%			xlabel={\bf mpACO computation times (s)},
%			ylabel={\bf Score},
%			]
%			\addplot+[
%			only marks,	
%			nodes near coords = {\numProcs{}},
%			visualization depends on = {value \thisrow{np} \as \numProcs},						
%			]
%			table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
%		\end{axis}
%		\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
%	\end{tikzpicture}
%		
%	\end{center}
%	\caption{ Scenario 1 performances with $1.2 \times volume$. }
%	\label{fig:scenario1data20}
%\end{figure}

@@

                      \cmidrule{2-10}		                       
                      &\multirow{2}{*}{ $mpShims$}& $f$               & {\bf 12.24} &  7.95      &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74} &   \\%
                      &                           & {\it run-time (s)}&  < 1        &  < 1       &    2        &   6         &    15       &   165       &   \\%

>>>>

\cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal. They used the {\it Scatter Search} metaheuristic as solution method.

@@

@@

The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$. It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.

@@

Below, we present the implementation for $SolveNode$, the new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.

@@

@@

@@

@@

\cmidrule{2-10}		                       
&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 12.24} &  7.95       &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74}  &   \\%
&                                                & {\it run-time (s)} &  < 1        &  < 1        &    2        &   6         &    15       &   165        &   \\%

\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & 0        & 0 & 13.38       &  0      & 0        &   \\%
&                                                & {\it run-time (s)} &  12         &   14        &  25         &  95         &    217      &  1719        &   \\%

\cmidrule{2-10}		                       
&\multirow{2}{*}{ $Shims$}                       & $f$                & 12.53       &  8.31       &  10.04      &  10.94      &   11.72     & 10.13        & \\%
&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         &\\%

\cmidrule{2-10}		                       
&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.59} & {\bf 10.49} &  16.82      & 18.46        & 22.19      & {\bf 0}  &   \\%
&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   10         &   27       &   165        &   \\%
